---
title: "Most of the productivity-pay gap is a unit conversion"
subtitle: "Output per hour is up 159 percent since 1973 and the production worker's real wage is up 3. Two thirds of the distance is deflator choice and the definition of a paycheck."
author: "Kyle Tong"
date: 2026-10-06
categories:
  - macro
  - data visualization
  - FRED
description: "The famous decoupling chart, taken apart into three pieces. The piece that is really about labor losing ground is the newest one, and it barely existed before 2000."
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(scales)

theme_set(theme_minimal(base_size = 12))

PAL <- c(productivity  = "grey20",
         comp_output   = "#0072B2",
         comp_consumer = "#009E73",
         wage_consumer = "#D55E00")

LAB <- c(productivity  = "output per hour",
         comp_output   = "compensation, output prices",
         comp_consumer = "compensation, consumer prices",
         wage_consumer = "production wage, consumer prices")

d <- read_csv(here("blog", "posts", "post4", "data", "derived",
                   "decomposition.csv"), show_col_types = FALSE)

last <- slice_tail(d, n = 1)
pt   <- function(x, dg = 0) formatC(x, format = "f", digits = dg)
sh   <- function(w) 100 * last[[w]] / last$gap_total

long <- d |>
  pivot_longer(all_of(names(PAL)), names_to = "series") |>
  mutate(series = factor(series, levels = names(PAL)))
```

Output per hour in the US nonfarm business sector is
`r pt(last$productivity - 100)` percent higher than it was in the first quarter of 1973.
The average hourly wage of a production worker, deflated by consumer prices, is
`r pt(last$wage_consumer - 100)` percent higher. Those two numbers are the usual evidence
that workers stopped sharing in what they produce.

```{r}
#| label: fig-gap
#| fig-cap: "Labor productivity and the production worker's real wage, 1973 Q1 = 100. The series start in 1964, where the wage data begins."
#| fig-width: 7
#| fig-height: 3.6
#| echo: false

long |>
  filter(series %in% c("productivity", "wage_consumer")) |>
  ggplot(aes(date, value, colour = series)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 100, linewidth = 0.3, colour = "grey70") +
  scale_colour_manual(values = PAL, labels = LAB, name = NULL) +
  labs(x = NULL, y = "index, 1973 Q1 = 100") +
  theme(legend.position = "top")
```

The two lines sit on top of each other until the mid-1970s and then separate, which is
where the story usually stops. But productivity and that wage are not measured in the same
units, and they do not refer to the same paycheck. Fixing each in turn closes most of the
gap.

## The deflator

Productivity is real output: what a sector produces, valued at the prices of the things it
produces. The wage above is deflated by the consumer price index, which measures what a
paycheck buys. A firm deciding whether an extra hour of labor is worth its cost compares
both to the price of its own output, so for that comparison the output deflator is the
consistent one.

The two price indices have not moved together. Housing, healthcare and education dominate
consumer spending and barely appear in business output prices, while computers and imported
goods pull the output deflator down. Swapping the consumer deflator for the output deflator,
holding the pay measure fixed, is worth `r pt(last$w_deflator)` index points.

## The paycheck

`AHETPI` covers production and nonsupervisory workers, roughly four fifths of private
employment, and it counts wages only. Compensation adds employer-paid health insurance and
retirement contributions, which is money the employer spends on labor whether or not it
shows up in a wage, and it covers everyone including managers. Moving from one to the other
is worth `r pt(last$w_pay_concept)` points.

This wedge is not purely bookkeeping. Part of it is benefits, and part of it is that pay
growth went to people above the production and nonsupervisory line. These series cannot
separate the two, and the second part is a real distributional change rather than a
measurement choice.

```{r}
#| label: fig-ladder
#| fig-cap: "The same gap, with each transformation added in turn. Each step up is one change of units or one change of pay concept."
#| fig-width: 7
#| fig-height: 4
#| echo: false

ggplot(long, aes(date, value, colour = series)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 100, linewidth = 0.3, colour = "grey70") +
  scale_colour_manual(values = PAL, labels = LAB, name = NULL) +
  labs(x = NULL, y = "index, 1973 Q1 = 100") +
  theme(legend.position = "top") +
  guides(colour = guide_legend(nrow = 2))
```

What survives both corrections is the distance between productivity and real compensation
measured in the same prices: `r pt(last$w_labor_share)` points, or
`r pt(sh("w_labor_share"))` percent of the original gap. That residual is labor's share of
output, and it is the only piece that means what the headline chart is usually taken to
mean.

## When each piece opened

```{r}
#| label: fig-wedges
#| fig-cap: "The gap in @fig-gap, split into three. The three lines add to the total at every date."
#| fig-width: 7
#| fig-height: 3.8
#| echo: false

d |>
  select(date, gap_total, w_pay_concept, w_deflator, w_labor_share) |>
  pivot_longer(-date) |>
  mutate(name = factor(name,
    levels = c("gap_total", "w_pay_concept", "w_deflator", "w_labor_share"),
    labels = c("total gap", "pay concept", "deflator", "labor share"))) |>
  ggplot(aes(date, value, colour = name, linewidth = name)) +
  geom_line() +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey70") +
  scale_colour_manual(values = c("grey70", "#009E73", "#0072B2", "#D55E00"),
                      name = NULL) +
  scale_linewidth_manual(values = c(1.2, 0.8, 0.8, 0.8), guide = "none") +
  labs(x = NULL, y = "index points of the gap") +
  theme(legend.position = "top")
```

The three pieces arrived in order. The pay-concept wedge opened first, in the late 1970s,
and had done most of its work by 2000 as employer health costs grew. The deflator wedge has
widened at a near-constant rate since the early 1980s. The labor-share wedge, the one that
actually describes workers getting a smaller slice, sat near zero for the first thirty
years and only crossed ten points in 2005. It is now `r pt(last$w_labor_share)`.

So the common summary, that pay stopped tracking productivity in the early 1970s, is
pointing at the right chart and the wrong decade. Labor's share of what it produced held up
through the 1980s and 1990s. It started falling around 2000, which is a narrower claim, a
more recent one, and a more useful one if the question is what changed.

## Caveats

Compensation here is an average, so it is pulled up by pay at the top; the distance between
average compensation and what a median worker earns is inequality rather than measurement,
and it is sitting inside the pay-concept wedge. Productivity is gross of depreciation, and a
net measure would close a few more points, since capital wears out faster than it used to.
The 1973 base quarter is a convention rather than a fact about the data.

Both scripts are in `R/`, with the series list and the construction in `README.md`. Data
from FRED, Federal Reserve Bank of St. Louis, accessed 6 October 2026.
