---
title: "Paying for wins"
subtitle: "Two web pages, 348 players, and the worst place in the NBA to spend money"
author: "Kyle Tong"
date: 2026-09-20
categories:
  - web scraping
  - rvest
  - sports economics
description: "Scraping Basketball-Reference salaries and production to find out which part of the market sells a win cheapest. It is not the part teams spend most in."
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(here)
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)

theme_set(theme_minimal(base_size = 12))

analysis <- read_csv(here("blog", "posts", "post2", "data", "derived",
                          "analysis_data.csv"), show_col_types = FALSE) |>
  mutate(band = factor(band, levels = c("Under $5M", "$5-15M",
                                        "$15-30M", "Over $30M")))

cost <- analysis |>
  group_by(band) |>
  summarise(players       = n(),
            median_salary = median(salary_m),
            cost_per_vorp = sum(salary) / sum(pmax(vorp, 0)) / 1e6)

price <- function(b) cost$cost_per_vorp[cost$band == b]
fmt   <- function(x, d = 1) formatC(x, format = "f", digits = d)
```

A team with $30 million of cap space this October can buy about
`r fmt(30 / price("Under $5M"))` points of VORP with it, or about
`r fmt(30 / price("$15-30M"))`, depending entirely on which part of the salary market it
shops in. VORP is Basketball-Reference's value over replacement player, and multiplying by
2.7 turns it into wins, so the gap between those two numbers is about
`r fmt(2.7 * (30 / price("Under $5M") - 30 / price("$15-30M")), 0)` wins for the same
money.

## Getting the data

Basketball-Reference publishes last season's player statistics and next season's contracts
as ordinary HTML tables. There is no API and no bulk download for them, which is what puts
this in scraping territory. Their `robots.txt` blocks game logs, splits and lineups but
leaves `/leagues/` and `/contracts/` open, and asks for a three second crawl delay, so the
acquisition script sends two requests four seconds apart, identifies itself in the user
agent, and saves the HTML to disk.

```{r}
#| label: scrape
#| eval: false

page <- read_html("https://www.basketball-reference.com/leagues/NBA_2026_advanced.html")
xml2::write_html(page, "data/raw/bbref_advanced_2026.html")

advanced <- page |>
  html_element("#advanced") |>
  html_table()
```

Saving the snapshot matters more than it looks. Every later step parses the local file, so
fixing a bug costs zero additional requests, and if the site redesigns its tables next
spring the analysis still runs against the page I actually saw.

Two things in the parsed tables would have quietly corrupted the result. Player names carry
bracketed footnote markers, so `Jokić[a]` and `Jokić` fail to join. And a player traded
mid-season gets one row per team plus a combined `2TM` row, all sharing a rank, so James
Harden appears three times unless the partial rows are dropped. Salaries arrive as strings
like `$62,587,158`, which `parse_number()` handles.

Because VORP is a fixed function of two other scraped columns,
$(\text{BPM} + 2) \times \text{MP} / (48 \times 82)$, recomputing it is a free test of the
whole parse. Across `r nrow(analysis)` players the largest disagreement is rounding, which
rules out shifted columns and numbers read as text.

## What production costs

Joining production to salary on player name matched 86 percent of contracts. The misses are
not spelling errors: they are 2026 draft picks with no NBA minutes and stars who missed all
of last season hurt, which is worth remembering when reading anything below about the top
of the market.

```{r}
#| label: fig-cost
#| fig-cap: "Total salary divided by total positive VORP, within each salary band. Players with at least 500 minutes in 2025-26, priced against their 2026-27 salary."
#| fig-width: 7
#| fig-height: 3.4
#| echo: false

ggplot(cost, aes(band, cost_per_vorp)) +
  geom_col(fill = "#2c7fb8", width = 0.62) +
  geom_text(aes(label = paste0("$", fmt(cost_per_vorp), "M")),
            vjust = -0.5, size = 3.2, colour = "grey25") +
  scale_y_continuous(labels = scales::label_dollar(suffix = "M"),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = "cost per point of VORP") +
  theme(panel.grid.major.x = element_blank())
```

Production gets more expensive as salaries rise, which is what competition for scarce
talent is supposed to do. The surprise is where the line stops rising. The most expensive
production in the league is not at the top: it is in the $15 to $30 million band, at
$`r fmt(price("$15-30M"))` million per point, above the $`r fmt(price("Over $30M"))`
million that players earning more than $30 million cost.

Two rules explain the shape. At the top, the collective bargaining agreement caps
individual salaries, so the best players in the league cannot be paid what they are worth.
At the bottom, minimum and rookie scale contracts are set by rule rather than by bidding,
and a young player who turns out to be good stays cheap for years. The $15 to $30 million
band has neither protection. It is priced by open negotiation for players good enough to be
paid but not good enough to be capped, and it is the one place where teams routinely pay
full price.

## What to do with that

For a team with cap space and no path to a superstar, the advice is to spend at the
extremes and stay out of the middle. Four players at $7 million who each produce half a
point of VORP return twice what one player at $28 million producing a single point does,
for the same money. The limit is the roster: 15 spots and 240 minutes a night, so a team
that already has useful cheap players cannot simply buy more of them, and improving a good
roster does eventually require paying for production per minute rather than per dollar.

Two caveats. VORP is a box score estimate and misses everything that does not end in a
counted event, and the salary is for next season while the production is from last season,
which will be wrong about exactly the players who change most.

Both scripts are in `R/`, the raw HTML snapshots and a provenance record in `data/raw/`,
and the tidy file behind the chart in `data/derived/`. Only `01_acquire.R` touches the
network. Data from Basketball-Reference, accessed 20 September 2026.
