---
title: "Participation fell, nobody stopped working"
subtitle: "US labor-force participation is down four points since 1990, and up inside every age group over 25"
author: "Kyle Tong"
date: 2026-09-28
categories:
  - labor market
  - data visualization
  - IPUMS CPS
description: "Participation is down four points since 1990. It is up inside every age group over 25. The difference is the age structure of the country."
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(scales)

theme_set(theme_minimal(base_size = 12))

AGE_COLOURS <- c("16-24" = "#D55E00", "25-54" = "#0072B2",
                 "55-64" = "#009E73", "65+"   = "#CC79A7")

DERIVED <- here("blog", "posts", "post3", "data", "derived")

lfp <- read_csv(file.path(DERIVED, "lfp_by_age.csv"), show_col_types = FALSE) |>
  mutate(age_group = factor(age_group, levels = names(AGE_COLOURS)))

gap <- read_csv(file.path(DERIVED, "weighting_gap.csv"), show_col_types = FALSE)

base <- lfp |>
  filter(year == 1990) |>
  select(age_group, lfp_1990 = lfp, share_1990 = share)

# Shift-share. "Rates only" lets participation move with age shares frozen at
# 1990; "shares only" lets the age structure move with 1990 participation.
agg <- lfp |>
  left_join(base, by = "age_group") |>
  group_by(year) |>
  summarise(actual      = sum(share * lfp),
            rates_only  = sum(share_1990 * lfp),
            shares_only = sum(share * lfp_1990),
            .groups     = "drop")

pick <- function(df, col, yr) 100 * df[[col]][df$year == yr]
pct  <- function(x, d = 1) formatC(x, format = "f", digits = d)
grp  <- function(g, yr) 100 * lfp$lfp[lfp$age_group == g & lfp$year == yr]
shr  <- function(g, yr) 100 * lfp$share[lfp$age_group == g & lfp$year == yr]
```

The share of Americans aged 16 and over who are working or looking for work fell from
`r pct(pick(agg, "actual", 1990))` percent in March 1990 to
`r pct(pick(agg, "actual", 2026))` percent in March 2026. Over the same 36 years,
participation rose in every age group above 25. Both statements come from the same
`r format(round(sum(lfp$n) / 1e6, 1), nsmall = 1)` million CPS records, and the distance
between them is the age structure of the country.

## The data

Every March Basic Monthly sample from IPUMS CPS between 1990 and 2026, with age,
labor-force status and the person weight `WTFINL`. The CPS is a sample, and its weights
exist to make that sample stand for the population, so every rate here is weighted:

```{r}
#| label: weighting
#| eval: false

adults |>
  group_by(year, age_group) |>
  summarise(lfp        = weighted.mean(in_lf, wtfinl),
            population = sum(wtfinl))
```

The universe is people 16 and over whose labor-force status is defined, which is what
`LABFORCE` in `(1, 2)` selects. Weighting is not a formality here. The unweighted
participation rate for March 2026 is `r pct(100 * gap$unweighted[gap$year == 2026])`
percent against a weighted `r pct(100 * gap$weighted[gap$year == 2026])`, because people
over 65 are `r pct(100 * sum(lfp$n[lfp$year == 2026 & lfp$age_group == "65+"]) / sum(lfp$n[lfp$year == 2026]))`
percent of the interviewed sample and `r pct(shr("65+", 2026))` percent of the population.
Older people answer the survey more often, and the CPS sample has shrunk from
`r format(sum(lfp$n[lfp$year == 1990]), big.mark = ",")` adults in 1990 to
`r format(sum(lfp$n[lfp$year == 2026]), big.mark = ",")`, so the correction has grown.

## Participation inside each age group

```{r}
#| label: fig-groups
#| fig-cap: "Weighted labor-force participation by age group, March of each year. Three of the four lines end above where they started."
#| fig-width: 7
#| fig-height: 3.8
#| echo: false

ggplot(lfp, aes(year, lfp, colour = age_group)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = AGE_COLOURS, name = NULL) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(x = NULL, y = "labor-force participation") +
  theme(legend.position = "top")
```

Workers aged 25 to 54 participate at `r pct(grp("25-54", 2026))` percent, against
`r pct(grp("25-54", 1990))` percent in 1990. The 55 to 64 group went from
`r pct(grp("55-64", 1990))` to `r pct(grp("55-64", 2026))` percent, and participation
after 65 rose from `r pct(grp("65+", 1990))` to `r pct(grp("65+", 2026))` percent. The one
group that pulled back is the youngest, down
`r pct(grp("16-24", 1990) - grp("16-24", 2026))` points, most of which is school
enrollment rather than idleness, though this extract cannot separate the two.

## The population behind the rates

```{r}
#| label: fig-shares
#| fig-cap: "Share of the population aged 16 and over, by age group. The 65+ group has gained eight points of the adult population since 1990."
#| fig-width: 7
#| fig-height: 3.8
#| echo: false

ggplot(lfp, aes(year, share, colour = age_group)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = AGE_COLOURS, name = NULL) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(x = NULL, y = "share of adult population") +
  theme(legend.position = "top")
```

People over 65 were `r pct(shr("65+", 1990))` percent of the adult population in 1990 and
are `r pct(shr("65+", 2026))` percent now. The prime-age group fell from
`r pct(shr("25-54", 1990))` to `r pct(shr("25-54", 2026))` percent. Since participation
after 65 is around a fifth and participation between 25 and 54 is above four fifths, that
reshuffling moves the national average on its own, without anyone changing their mind
about work.

## Splitting the two apart

```{r}
#| label: fig-decomp
#| fig-cap: "Aggregate participation, and two counterfactuals. The dashed line holds age shares at their 1990 values; the dotted line holds each group's participation rate at its 1990 value."
#| fig-width: 7
#| fig-height: 4
#| echo: false

agg |>
  pivot_longer(-year) |>
  mutate(name = factor(name, levels = c("actual", "rates_only", "shares_only"),
                       labels = c("actual",
                                  "1990 age shares, actual rates",
                                  "actual age shares, 1990 rates"))) |>
  ggplot(aes(year, value, colour = name, linetype = name)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("grey20", "#0072B2", "#D55E00"), name = NULL) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted"), name = NULL) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(x = NULL, y = "participation rate, 16 and over") +
  theme(legend.position = "top", legend.text = element_text(size = 9)) +
  guides(colour = guide_legend(nrow = 2), linetype = guide_legend(nrow = 2))
```

Freeze the age structure at its 1990 shares and let participation move as it actually did:
the rate ends at `r pct(pick(agg, "rates_only", 2026))` percent, slightly above where it
started. Freeze participation instead and let only the age structure move: the rate falls
to `r pct(pick(agg, "shares_only", 2026))` percent. The entire decline in the headline
number, and more, is demographics.

The actual line sits above the demographics-only line, which is the interesting residual.
The groups gaining population share are the same ones raising their participation, so
older Americans working longer has offset about
`r pct(pick(agg, "actual", 2026) - pick(agg, "shares_only", 2026))` points of what aging
would otherwise have taken out.

## Caveats

Each point is one March, so 2021 carries the pandemic and single months are noisier than
annual averages. The CPS reweights to new population controls after each census, which
puts small breaks in the series around 2000, 2010 and 2020. And a shift-share split like
this one describes an accounting identity, not a cause: it says how much of the change
sits in the age structure, not why the age structure moved or why 60-year-olds now work.

Both scripts are in `R/`. IPUMS licensing does not allow redistributing microdata, so
`data/raw/` stays out of the repository and only the aggregated table in `data/derived/`
is committed. Data from IPUMS CPS, accessed 28 September 2026.
